---
title: "EpiR"
author: "grienne"
date: "February 22, 2019"
output: pdf_document
---

```{r setup, echo=FALSE}
knitr::opts_chunk$set(tidy.opts=list(width.cutoff=60),tidy=TRUE, echo = TRUE)

```


##Welcome to RMarkdown Documents!

RMarkdown documents are how many people produce files that are shareable for efficient collaborative work!

Please note the annotations before the text by the *'#'* symbol.

Quick Note: Running Code

1. Highlight the code then click ctrl+enter

2. There is a green forward arrow in the upper right of these highlighted areas with code in them (these are called chunks btw!). Click that arrow and it will run all the code in sequential order!


##Packages


The code below checks your R to make sure you have the packages needed and if you don't it downloads and installs them!

**IMPORTANT!** 

In this template I am using a data set in the *MASS* package. However, you will **NOT** need this package. When you are using this code, **DELETE** that package!

Your packages statement should look like this:

*packages = c("epiR", "tidyverse", "survival", "readr")*

```{r Packages, message=FALSE}

#specify the packages of interest
packages = c("epiR", "tidyverse", "survival", "readr", "MASS")

#use this function to check if each package is on the local machine
#if a package is installed, it will be loaded
#if any are not, the missing package(s) will be installed and loaded
package.check <- lapply(packages, FUN = function(x) {
  if (!require(x, character.only = TRUE)) {
    install.packages(x, dependencies = TRUE)
    library(x, character.only = TRUE)
  }
})
```


###Loading Data


Loading Data in R can be intensive, but luckily RStudio makes this fairly easy!

The link below is a website that will guide you through that process in RStudio.

Loading Data: [link]https://support.rstudio.com/hc/en-us/articles/218611977-Importing-Data-with-RStudio 

*This is a sample dataset that is located in the MASS Package. You will be working with ARIC. Some additional code will be provided for you at a later date*

When you load a data file, I encourage you to rename it to something clear and short!
The code below does 2 things:
    1. I take the loaded dataset (named birthwt) and rename it to dat1 'dat1 <- birthwt'
        *Note: You can rename it however you want*
    2. I take a quick look at the dataset to make sure it looks okay 'head(dat1)'

```{r Data}

dat1 <- birthwt; head(dat1)

#R is case sensitive, typically it is good practice 
#to make everything lowercase so you don't get errors because of case heterogeneity.
names(dat1) <- tolower(names(dat1))

```


##---DATA MANAGEMENT Part 1 --- Subsetting Data

###Subsetting Data
  There are a lot of ways to subset data! 
  
*The **detach** command removes the MASS package so I can use the **select** command.* 
  *You will not need the detach command code line as you will not have MASS loaded for your work.* 
  *However, if you want to mess around with this template using the birthweight* 
    *data set then leave this code as is*
  
```{r, tidy.opts=list(width.cutoff=60)}
#IMPORTANT! Delete the detach code line if you do not need it!

detach("package:MASS", unload = TRUE)


#Subsetting Tips
##I suggest creating a dataset when subsetting, sometimes you make unfixable or                                         annoying mistakes, it is easier at times to go back to your original dataset and try again. 

#---------------------------------------------------------

###This code works by taking the original data (dat1) then selecting the columns                                       you want instead to create another dataset. In the example below I wanted the following variables: low, lwt, and race

dat2 <- dat1 %>%
        select(low, lwt, race)

```


##---DATA MANAGEMENT PART 2--- Variable Management

###Variable Attribute Control


We need to talk about variable attributes. Every statistical software and database program manages data by controlling managing variable "attributes."

Think of using the Excel software. If you want to change the type of data or format of the data you can 'format cells' to change the column data type to dates, integers, text, etc.  

R controls data by identifying variable attributes. Attribute control is crucial to effective code-writing.  

For this class you need to know that categorical variables should be *factors* & continuous data should be *integers*
```{r}

#The str command produces an output that tells you the attribute of a variable. 'Int' means integer, Facotr means factor
#When you import data, you may see 'str' or 'chr', this means string and character respectively. 

str(dat1)

#Turn the variables for 2x2 contingency tables into factors

dat1$low <- as.factor(dat1$low)
dat1$smoke <- as.factor(dat1$smoke)
dat1$race <- as.factor(dat1$race)

#check to make sure you changed it!
str(dat1)

```



###Renaming Data

I don't like working with 1's and 0's if the data is categorical. R doesn't really care as long as the variable attribute is correct, but often to label my data to keep my code output simple and easily readable!

This might look intimidating but I promise it's fairly simple!

This uses an *'ifelse'* statement to recode the numbers (now factors) into words. We will break this code down to the components

Note: *R really cares about the parentheses so make sure every parentheses has it's partner!*
```{r}
#1.'dat1$race <-  
    #tells R which dataset and which column (race) you are working on. 
    #with(dat1, --- this is the beginning of the code for changing, it must be ordered like this!

#2. if_else(column name == outcome, 'What you want it to be if true',                                                        'what you want it to be if false')
    #Now, I am changing multiple outcome possibilities so it says:
      #If column 'race' outcome equals 1, that means 'black', but if it equals 2....etc
      #The last one 'Other' is the 'what...if false' statement. 
      #YOU ALWAYS NEED something here, so just keep in mind what                                                                you want your column to say if the other values arent there, here                                                        I knew that if it wasn't 1-3 the data should be coded as 'other'
    
dat1$race <- with(dat1,
                        if_else(race == 1, "Black",
                            ifelse(race == 2, "White",
                               ifelse(race == 3, "Non-White Hispanic", "Other"
                                        )
                                       )
                                      )
                                     )

```




###Filtering Data


Often times you will want to filter your data! Data filtering can be easy or complex, it really depends on your needs, for this course it should be pretty easy!

I am going to show you how to filter categorical data & continuous data; there are lots of different ways to do this. R uses basic LOGICAL OPERATORS just like excel for the most part!

I am going to take the dat2 we created above and use that for my example! 

There are many logical operators, I present a few below, but you may need others [link]https://www.datamentor.io/r-programming/operator/ 

Note: *As a personal preference, when filtering I usually don't create new datasets, but feel free to create new ones if it is easier for you, just change the name on the left side of the code*
```{r}

#Same if_else pattern as above
##I am turning the race category so that it is easier for me to filter. This is really                                  just preference, I do not like working with 1's and 0's. 
dat2$race <- with(dat2,
                        if_else(race == 1, "Black",
                            ifelse(race == 2, "White",
                               ifelse(race == 3, "Non-White Hispanic", "Other"
                                        )
                                       )
                                      )
                                     )

#------------------------------------

dat2 <- dat2 %>%
          #categorical variables must be bounded by the quotation marks
    
          filter(race == "Black") %>%
        
          #Continuous variables do not need 'quotation' marks, just basic mathematical operators
          #Below I said all lwt greater than 107
          
          filter(lwt > 107) %>%
  
          #Here I tell R to select lwt output that is greater than/equal to 130, but less/equal to 200
  
          filter(lwt >= 130 & lwt <= 200)

          #Note How I did 2 filters at once! This is done by PIPING! I won't go into it,                                            but basically at the end of the first filter, I did another %>%,                                                          that lets R know to run another process. To end a pipe just don't add a %>% at the end!
          
          #Tip: I could actually have did the selection and filtering all                                                             at the same time using this process when I am working, I tend                                                             to do that, but I encourage you instead to split these things up.                                                         Until you're used to it, it makes it easier to                                                                            identify possible problems and remember what you were doing

```


###Descriptive Statistics

```{r}

#The quick descriptive statistics analysis
summary(dat1)

#If you want to do a summary for only one variable

summary(dat1$age)
#When you use the '$' sign in front of the named data set, a box will appear                                              with your variables, you can select it there or just continue typing the variable name

```

#----------------------------------------------------------------

##Generate Contingency Tables

Order of the code input matters! So I suggest identifying your outcome/exposure variables first. 
This way you know exactly where to put them in the code below!

Note: *Some of the output from these tables won't be used until later in the course! But it all generates so just ignore what you don't need for now*

###---Ratios---

Calculations! Finally, it's time to do some actual stats!

Note: *The command: **conf.level =**, The default in R is 95% and this is what you will assume for this course. However, this can be changed to whatever you want *


**IMPORTANT!**: When you finish the epi.2by2 command, you will see output in the console box. However, under the chunk RMarkdown will produce 2 selectable frames. A data.frame box and an R Console Box. The data.frame has the incidence risk, odds ratios, and the incidence rate (when appropriate)

```{r}

## Generate the 2 by 2 table. Exposure (rows) = smoke. Outcome (columns) = low. 

## Note order of variables; DNN command places titles IN THE FINAL OUTPUT
tab1 <- table(dat1$smoke, dat1$low, dnn = c("Smoke", "Low BW"))

#Print, tells R to actually show the table it made!
print(tab1)


#--------------------------------------------

#Units = 100, tells R in what unit count, so this is per 100 of whatever your primary unit is
t1r <- epi.2by2(dat = tab1, method = "cohort.count",
         conf.level = 0.95, units = 100, homogeneity = "breslow.day",
         outcome = "as.columns")


print(t1r$res$OR.strata.cfield)


#The outcomes from this table with the 95% confidence intervals
#Incident Risk Ratio
#Odds Ratio
#Attributable Risk
#Attr Risk in Pop
#Attrib. Fraction in exposed
#Attrib. Fraction in Pop.

#---------------------------------------------------------------

#Sample Interpretation
## Odds ratio:
## The odds of having a low birth weight child for smokers is 2.02
## (95% CI 1.08 to 3.78) times greater than the odds of having
## a low birth weight child for non-smokers.

```



##---Stratification---


The code below stratifies by race. 

Note: *Note the output order of *print(tab2)* when you generate the OR per strata the output will be in the same order*

```{r}

## Now stratify by race: 
## Note added a factor in the code. This is telling R to stratify by Race 
## Remember, order matters!!


tab2 <- table(dat1$smoke, dat1$low, dat1$race,
              dnn = c("Smoke", "Low BW", "Race"))
print(tab2)



## Compute the crude odds ratio, the Mantel-Haenszel adjusted odds ratio
## and other measures of association:
rval <- epi.2by2(dat = tab2, method = "case.control",
                 conf.level = 0.95, units = 100, homogeneity = "breslow.day",
                 outcome = "as.columns")

print(rval)

#Generate The OR per strata with confidence intervals
rval$res$OR.strata.wald


##Sample Interpretation:
## After accounting for the confounding effect of race, the odds of
## having a low birth weight child for smokers is 3.09 (95% CI 1.49 to 6.39)
## times that of non-smokers.

```


#----------------------------------------------------------------


##---Rates---
```{r}
#Now we will compute INCIDENCE RATES


#Essentially you need 2 columns; cases and the person-time (c7_futime)
#In the ARIC data set both of these columns are provided
#As you continue on, you will learn to identify those columns!

dat <- as.table(matrix(c(136,22050,1709,127650), nrow = 2, byrow = TRUE))

#note structure: first column is cases, second is person time
#While this data set is created, remember as above, you will create a tab, and order matters!
print(dat)

rval <- epi.2by2(dat = dat, method = "cohort.time", conf.level = 0.95,
                 units = 1000, homogeneity = "breslow.day", outcome = "as.columns")

#The outcomes from this table with the 95% confidence intervals
#Incident Rate Ratio
#Attrib Rate
#Attrib Rate in Population
#Attr Frac in Exposed(%)
#Attrib. Fraction in Pop.

print(rval)

#Incidence Rate
#The code above generates all of the information needed
#The next code line pulls out the incidence rate specifically
summary(rval)$ARate.strata.wald

##Interpretation
## The incidence rate of cancer was 7.22 cases per 1000 person-years less in the
## blind, compared with those who were not blind but had severe visual impairment
## (90% CI 6.00 to 8.43 cases per 1000 person-years).


#--------------------


#The code below does a rate ratio of the two groups.
round(summary(rval)$IRR.strata.wald, digits = 2)

#Interpretation
## The incidence rate of cancer in the blind group was less than half that of the
## comparison group (incidence rate ratio 0.46, 90% CI 0.38 to 0.55).
```


##---Complete Rows---

Sometimes you need to evaluate if your data is complete. R has a useful way of telling you how many of your observations are missing values (NA)

```{r}
#I create an object that evaluates the number of complete cases in a data set
ok <- complete.cases(dat1)

#I sum the number of incomplete observations
sum(!ok)
```