---
title: "Project Part 2"
author: "grienne"
date: "March 26, 2019"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

###Packages

```{r Packages, message=FALSE, eval=FALSE}

#specify the packages of interest
packages = c("epiR", "tidyverse", "survival", "readr", "tableone", "purrr", "lattice")

#use this function to check if each package is on the local machine
#if a package is installed, it will be loaded
#if any are not, the missing package(s) will be installed and loaded
package.check <- lapply(packages, FUN = function(x) {
  if (!require(x, character.only = TRUE)) {
    install.packages(x, dependencies = TRUE)
    library(x, character.only = TRUE)
  }
})


```

###Loading Data


Loading Data in R can be intensive, but luckily RStudio makes this fairly easy!

The link below is a website that will guide you through that process in RStudio.

Loading Data: [link]https://support.rstudio.com/hc/en-us/articles/218611977-Importing-Data-with-RStudio 

When you load a data file, I encourage you to rename it to something clear and short!
The code below does 2 things:
    1. I take the loaded dataset rename it to dat1 'dat1 <- academic_dataset_arc.csv'
        *Note: You can rename it however you want*
    2. I take a quick look at the dataset to make sure it looks okay 'head(dat1)'


```{r eval=FALSE}

dat1 <- read_csv("academic_dataset_aric.csv"); head(dat1)
names(dat1) <- tolower(names(dat1))

#I removed the enddate and dated11 column as it lengthened the output significantly
drop.cols <- c("c7_enddate", "dated11")
dat2 <- dat1 %>% 
        select(-one_of(drop.cols))

dat3 <- dat2 %>%
        select(v1age01, bmi01, gender, racegrp, hyptmd01, diabts03, hdlsiu02, ldlsiu02, tchsiu01, trgsiu01, fast0802, cigt01, prvchd05, center, c7_inc_by11, c7_futime)

```



####Question 1 - Descriptive Statistics

The code below generates descriptive statistics for the dataset we created above, including means, medians, range, and modes. 

Also identifies number of NA observations.
```{r eval=FALSE}

summary(dat3)

```

####Question 1a - Race Removal
```{r eval=FALSE}

#removed races A & I
dat4<- dat3 %>%
        filter(racegrp != "A" & racegrp != "I")

#turning Racegrp into a factor (i.e. making it categorical)
dat4$racegrp <- as.factor(dat4$racegrp)

#We have removed A&I so when we do summary of dat and just check the Racegrp there should only be Black and White
summary(dat4$racegrp)

#Making Center a Factor
dat4$center <- as.factor(dat4$center)

#Summarize data by Center
#This produces descriptive statistics by Center. Including Race Counts
dat4 %>%
    split(.$center) %>%
      map(summary)

#remove D & B black participants 


#Code Tip: Sometimes you have to effect change on an existing dataset. However, if you aren't sure the code will work, I will often create a subset of it
dat_r <- dat4[!(dat4$center == "B" & dat4$racegrp == "B"),]
dat_r <- dat4[!(dat4$center == "D" & dat4$racegrp == "B"),]

#Removed 6 blacks from Center B & 13 from center D
#I reassign dat_r to dat4 once I have verified I am okay with the outcome

dat4 <- dat_r

summary(dat4)

#Histogram is made using the below command. 
  #Tips
    #To make a histogram dat3$ and select the variable of interest
    #If you want to make a histogram by category the first variable is the continuous variable, the second is the               categorical variable
histogram(~ dat4$hdlsiu02|dat4$racegrp)

```


####Question 1 - Race Center

Here I am creating the race center variable
```{r eval=FALSE}

#Create Race Center
dat5 <- dat4 %>%
            mutate(race.center = ifelse(racegrp == "W" & center == "A", 1,
                          ifelse(racegrp == "B" & center == "A", 2,
                            ifelse(racegrp == "W" & center == "B", 3,
                               ifelse(racegrp == "W" & center == "D", 4,
                                  ifelse(racegrp == "B" & center == "C", 5, NA)
                                       )))))

#Factor
#Turning multiple variables into factors/categories 
varstofactor <- c("gender", "racegrp", "cigt01", "prvchd05", "diabts03", "c7_inc_by11")
dat5[varstofactor] <- lapply(dat5[varstofactor], factor)

#Summary
#I am summarizing dat 5, then summarizing specific specific variables from data 5
dat5$race.center <- as.factor(dat5$race.center)
summary(dat5)
summary(dat5$hyptmd01)
summary(dat5$race.center)


rm(dat_r); rm(dat1); rm(dat2); rm(dat3); rm(dat4)
```


##Epi 2x2

```{r}
summary(dat5)

summary(dat5$c7_inc_by11)

```

Question 1 - Incidence Rates

##---Incidence Rate Overall
```{r}

##Overall
summary(dat5$c7_inc_by11)
#4031 - 0
#639 - 1

tot <- dat5 %>%
    group_by(c7_inc_by11) %>%
      summarise(Total = sum(c7_futime))

print(tot)

#0	30122969			
#1	2651505	


dat <- as.table(matrix(c(639,2651505, 4031,30122969), nrow = 2, byrow = TRUE))
rownames(dat) <- c("Exp + ", "Exp -"); colnames(dat) <- c("Disease +", "Person-Time"); dat


rval <- epi.2by2(dat = dat, method = "cohort.time", conf.level = 0.95,
                 units = 1000, homogeneity = "breslow.day", outcome = "as.columns")

print(rval)

#Incidence Rate
#The code above generates all of the information needed
#The next code line pulls out the incidence rate specifically
summary(rval)$ARate.strata.wald

#--------------------

round(summary(rval)$IRR.strata.wald, digits = 2)


```

##Incidence Rate by Gender

```{r}


tab1 <- table(dat5$c7_inc_by11, dat5$gender, dnn = c("Myocardial Infarction", "Gender"))
print(tab1)

totg <- dat5 %>%
    group_by(c7_inc_by11, gender) %>%
      summarise(Total = sum(c7_futime))

print(totg)

#0	F	18186455	2368
#1	F	1091282		242
#0	M	11936514	1663	
#1	M	1560223	  397

#---Female
#
rate_female <- as.table(matrix(c(242, 1091282 ,2368, 18186455 ), nrow = 2, byrow = TRUE))
rownames(rate_female) <- c("Myo + ", "Myo - "); colnames(rate_female) <- c("Disease +", "Person-Time"); rate_female


rval_female <- epi.2by2(dat = rate_female, method = "cohort.time", conf.level = 0.95,
                      units = 1000, homogeneity = "breslow.day", outcome = "as.columns")

print(rval_female)

#-----Male

rate_male <- as.table(matrix(c(397, 1560223 ,1663, 11936514), nrow = 2, byrow = TRUE))
rownames(rate_male) <- c("Myo + ", "Myo - "); colnames(rate_male) <- c("Disease +", "Person-Time"); rate_male


rval_male <- epi.2by2(dat = rate_male, method = "cohort.time", conf.level = 0.95,
                      units = 1000, homogeneity = "breslow.day", outcome = "as.columns")

print(rval_male)

#----Rate Difference

1.83 - 1.70

```


##Incidence Rate by Race

```{r}


tabR <- table(dat5$c7_inc_by11, dat5$racegrp, dnn = c("Myocardial Infarction", "Race"))
print(tabR)

totR <- dat5 %>%
    group_by(c7_inc_by11, racegrp) %>%
      summarise(Total = sum(c7_futime))

print(totR)

#0	B	8205785	  1129
#1	B	748864		178	
#0	W	21917184	2902	
#1	W	1902641	  461

#---Black

rate_black <- as.table(matrix(c(178, 748864, 1129, 8205785), nrow = 2, byrow = TRUE))
rownames(rate_black) <- c("Myo +", "Myo -"); colnames(rate_black) <- c("Disease +", "Person-Time"); rate_black


rval_black <- epi.2by2(dat = rate_black, method = "cohort.time", conf.level = 0.95,
                      units = 1000, homogeneity = "breslow.day", outcome = "as.columns")

print(rate_black)


#----White

rate_white <- as.table(matrix(c(461, 1902641, 2902, 21917184), nrow = 2, byrow = TRUE))
rownames(rate_white) <- c("Myo +", "Myo -"); colnames(rate_white) <- c("Disease +", "Person-Time"); rate_white


rval_white <- epi.2by2(dat = rate_white, method = "cohort.time", conf.level = 0.95,
                      units = 1000, homogeneity = "breslow.day", outcome = "as.columns")

print(rate_white)

#---Rate Difference

#Input

```

##Incidence Rate by Education Level --- FIX -- Which is the Education Variable

```{r}


tab1 <- table(dat5$c7_inc_by11, dat5$racegrp, dnn = c("Myocardial Infarction", "Gender"))
print(tab1)

tot <- dat5 %>%
    group_by(c7_inc_by11, gender) %>%
      summarise(Total = sum(c7_futime))

print(tot)

#0	F	18186455		
#0	M	11936514		
#1	F	1091282		
#1	M	1560223	


rate_female <- as.table(matrix(c(242, 1091282 ,397,1560223	), nrow = 2, byrow = TRUE))
rownames(rate_female) <- c("Female ", "Male"); colnames(rate_female) <- c("Disease +", "Person-Time"); rate_female


rval_gender <- epi.2by2(dat = rate_female, method = "cohort.time", conf.level = 0.95,
                      units = 1000, homogeneity = "breslow.day", outcome = "as.columns")

print(rval_gender)
```


##Incidence Rate by Diabetes Status

```{r}


tabD <- table(dat5$c7_inc_by11, dat5$diabts03, dnn = c("Myocardial Infarction", "Diabetes"))
print(tabD)

totD <- dat5 %>%
    group_by(c7_inc_by11, diabts03) %>%
      summarise(Total = sum(c7_futime))

print(totD)

#0	0	27361132 3606		
#0	1	2489190	 387
#1	D0	2049250 478		
#1	D1	574243 155	

#--Diabetes
dat_diab_p <- as.table(matrix(c(155, 574243 , 478, 2049250	), nrow = 2, byrow = TRUE))
rownames(dat_diab_p) <- c("Myo + ", "Myo -"); colnames(dat_diab_p) <- c("Disease +", "Person-Time"); dat_diab_p


rval_diab_p <- epi.2by2(dat = dat_diab_p, method = "cohort.time", conf.level = 0.95,
                      units = 1000, homogeneity = "breslow.day", outcome = "as.columns")

print(rval_diab_p)


#--No Diabetes
dat_diab_n <- as.table(matrix(c(155, 574243 , 478, 2049250	), nrow = 2, byrow = TRUE))
rownames(dat_diab_n) <- c("Myo + ", "Myo -"); colnames(dat_diab_n) <- c("Disease -", "Person-Time"); dat_diab_n


rval_diab_n <- epi.2by2(dat = dat_diab_n, method = "cohort.time", conf.level = 0.95,
                      units = 1000, homogeneity = "breslow.day", outcome = "as.columns")

print(rval_diab_n)

#---Rate Difference

```


#Incidence Ratios


##Incidence Ratio by Gender
```{r}


tab_gender <- table(dat5$c7_inc_by11, dat5$gender, dnn = c("Myocardial", "Gender"))
print(tab_gender)


## Compute the crude odds ratio, the Mantel-Haenszel adjusted odds ratio
## and other measures of association:
eval_gender <- epi.2by2(dat = tab_gender, method = "case.control",
                 conf.level = 0.95, units = 100, homogeneity = "breslow.day",
                 outcome = "as.columns")

print(eval_gender)

#Generate The OR per strata with confidence intervals
eval_gender$res$OR.strata.wald

```


##Incidence Ratio by Race
```{r}


tab_race <- table(dat5$c7_inc_by11, dat5$racegrp, dnn = c("Myocardial", "Race"))
print(tab_race)


## Compute the crude odds ratio, the Mantel-Haenszel adjusted odds ratio
## and other measures of association:
eval_race <- epi.2by2(dat = tab_race, method = "case.control",
                 conf.level = 0.95, units = 100, homogeneity = "breslow.day",
                 outcome = "as.columns")

print(eval_race)

#Generate The OR per strata with confidence intervals
eval_race$res$OR.strata.wald


```


##Incidence Ratio by Education --- FIX
```{r}


tab_gender <- table(dat5$c7_inc_by11, dat5$gender, dnn = c("Myocardial", "Gender"))
print(tab_gender)


## Compute the crude odds ratio, the Mantel-Haenszel adjusted odds ratio
## and other measures of association:
eval_gender <- epi.2by2(dat = tab_gender, method = "case.control",
                 conf.level = 0.95, units = 100, homogeneity = "breslow.day",
                 outcome = "as.columns")

print(eval_gender)

#Generate The OR per strata with confidence intervals
rval$res$OR.strata.wald


```


##Incidence Ratio by Diabetes
```{r}


tab_diab <- table(dat5$c7_inc_by11, dat5$diabts03, dnn = c("Myocardial", "Diabetes"))
print(tab_gender)


## Compute the crude odds ratio, the Mantel-Haenszel adjusted odds ratio
## and other measures of association:
eval_diab <- epi.2by2(dat = tab_diab, method = "case.control",
                 conf.level = 0.95, units = 100, homogeneity = "breslow.day",
                 outcome = "as.columns")

print(eval_diab)

#Generate The OR per strata with confidence intervals
eval_diab$res$OR.strata.wald


```





