---
title: "Project 3"
author: "grienne"
date: "March 28, 2019"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

###Packages

```{r Packages, message=FALSE, eval=FALSE}

#specify the packages of interest
packages = c("epiR", "tidyverse", "survival", "readr", "tableone", "purrr", "lattice")

#use this function to check if each package is on the local machine
#if a package is installed, it will be loaded
#if any are not, the missing package(s) will be installed and loaded
package.check <- lapply(packages, FUN = function(x) {
  if (!require(x, character.only = TRUE)) {
    install.packages(x, dependencies = TRUE)
    library(x, character.only = TRUE)
  }
})


```

###Loading Data


Loading Data in R can be intensive, but luckily RStudio makes this fairly easy!

The link below is a website that will guide you through that process in RStudio.

Loading Data: [link]https://support.rstudio.com/hc/en-us/articles/218611977-Importing-Data-with-RStudio 

When you load a data file, I encourage you to rename it to something clear and short!
The code below does 2 things:
    1. I take the loaded dataset rename it to dat1 'dat1 <- academic_dataset_arc.csv'
        *Note: You can rename it however you want*
    2. I take a quick look at the dataset to make sure it looks okay 'head(dat1)'


```{r eval=FALSE}

dat1 <- read_csv("academic_dataset_aric.csv"); head(dat1)
names(dat1) <- tolower(names(dat1))

#I removed the enddate and dated11 column as it lengthened the output significantly
drop.cols <- c("c7_enddate", "dated11")
dat2 <- dat1 %>% 
        select(-one_of(drop.cols))

dat3 <- dat2 %>%
        select(v1age01, bmi01, gender, racegrp, hyptmd01, diabts03, hdlsiu02, ldlsiu02, tchsiu01, trgsiu01, fast0802, cigt01, prvchd05, center, c7_inc_by11, c7_futime)

```



####Descriptive Statistics

The code below generates descriptive statistics for the dataset we created above, including means, medians, range, and modes. 

Also identifies number of NA observations.
```{r eval=FALSE}

summary(dat3)

```

####Race Removal
```{r eval=FALSE}

#removed races A & I
dat4<- dat3 %>%
        filter(racegrp != "A" & racegrp != "I")

#turning Racegrp into a factor (i.e. making it categorical)
dat4$racegrp <- as.factor(dat4$racegrp)

#We have removed A&I so when we do summary of dat and just check the Racegrp there should only be Black and White
summary(dat4$racegrp)

#Making Center a Factor
dat4$center <- as.factor(dat4$center)

#Summarize data by Center
#This produces descriptive statistics by Center. Including Race Counts
dat4 %>%
    split(.$center) %>%
      map(summary)

#remove D & B black participants 


#Code Tip: Sometimes you have to effect change on an existing dataset. However, if you aren't sure the code will work, I will often create a subset of it
dat_r <- dat4[!(dat4$center == "B" & dat4$racegrp == "B"),]
dat_r <- dat4[!(dat4$center == "D" & dat4$racegrp == "B"),]

#Removed 6 blacks from Center B & 13 from center D
#I reassign dat_r to dat4 once I have verified I am okay with the outcome

dat4 <- dat_r

summary(dat4)

#Histogram is made using the below command. 
  #Tips
    #To make a histogram dat3$ and select the variable of interest
    #If you want to make a histogram by category the first variable is the continuous variable, the second is the               categorical variable
histogram(~ dat4$hdlsiu02|dat4$racegrp)

```


####Race Center

Here I am creating the race center variable
```{r eval=FALSE}

#Create Race Center
dat5 <- dat4 %>%
            mutate(race.center = ifelse(racegrp == "W" & center == "A", 1,
                          ifelse(racegrp == "B" & center == "A", 2,
                            ifelse(racegrp == "W" & center == "B", 3,
                               ifelse(racegrp == "W" & center == "D", 4,
                                  ifelse(racegrp == "B" & center == "C", 5, NA)
                                       )))))

#Factor
#Turning multiple variables into factors/categories 
varstofactor <- c("gender", "racegrp", "cigt01", "prvchd05", "diabts03", "c7_inc_by11")
dat5[varstofactor] <- lapply(dat5[varstofactor], factor)

#Summary
#I am summarizing dat 5, then summarizing specific specific variables from data 5
dat5$race.center <- as.factor(dat5$race.center)
summary(dat5)
summary(dat5$hyptmd01)
summary(dat5$race.center)


rm(dat_r); rm(dat1); rm(dat2); rm(dat3); rm(dat4)
```




#Logistic Regression

```{r}

model_1 <- glm(dat5$diabts03 ~ dat5$c7_futime , family= "binomial", data = dat5)
summary(model_1)

```


```{r}

```


```{r}

model_3 <- glm(dat5$diabts03 ~ dat5$c7_futime , family= "binomial", data = dat5)
summary(model_3)

```

